Automatically Splitting a Two-Stage Lambda Calculus

### 2016-01-08 paper/intro.tex
mostly compression
small edits

\input{figures/quickselect}
Consider a function $F$ which takes an $x$ and $y_1,\dots,y_m$, and computes
$f(x,y_i)$ for each $y_i$:
\begin{lstlisting}
fun @$F(x, y_1, y_2 \ldots, y_m)$@ = 
  @$f(x, y_1);  f(x, y_2); \ldots; f(x, y_m)$@
\end{lstlisting}
%
Observe that this would be wasteful if $f(x,y_i)$ does a significant 
amount of work that does not depend on $y_i$.
In such a case, it would be advantageous to find the computations in
$f$ that depend only on $x$ and then stage execution of $f$ so they are
performed only once at the beginning of $F$.

>>>>

\input{figures/quickselect}
Consider a function $F$ which, given $x, y_1,\dots,y_m$, computes
$f(x,y_i)$ for each $y_i$:
\begin{lstlisting}
fun @$F(x, y_1, y_2 \ldots, y_m)$@ = 
  @$f(x, y_1);  f(x, y_2); \ldots; f(x, y_m)$@
\end{lstlisting}
%
Observe that this implementation would be wasteful if $f(x,y_i)$ does a significant 
amount of work that does not depend on $y_i$.
In such a case, it would be advantageous to find the computations in
$f$ that depend only on $x$ and then stage execution of $f$ so they are
performed only once at the beginning of $F$.

### 2016-01-08 paper/overview.tex
small edits

We can attempt to improve on this algorithm by factoring out computations shared
between these calls to \texttt{qs}, 
namely the parts that depend only on \texttt{l}.
A first cut at a solution might be to first sort \texttt{l} into
a sorted list \texttt{s}, at cost $\Theta(n \log n)$, and then perform lookups in \texttt{s}.
Unfortunately, lookups in \texttt{s} are $\Theta(n)$ on average, 
since \texttt{s} is a list.
Thus, the total cost is $\Theta(n \log n + mn)$, which is no improvement.

A more sophisticated solution is to construct a binary search
tree out of \texttt{l}, again at cost $\Theta(n\log{n})$, and then simply look for the \texttt{k}th
leftmost element of that tree. 
Doing lookups efficiently, in $\Theta(\log{n})$ time, 
requires one more innovation---augmenting the tree 
by storing at each node the size of its left subtree. 
Thus in total this approach has expected runtime $\Theta(n\log{n} + m\log{n})$.

Is this method, wherein we precompute a data structure, better than the direct $\Theta(mn)$ method from above? 
The answer depends on the relationship between the number of lookups $m$ and the size of the list $n$.
If $m$ is constant, then the direct method is superior for sufficiently large $n$.
This is because the precomputed method does unnecessary work 
sorting parts of the list where there are no query points.
However, if the number of queries grows with the size of the list, 
specficially in $\omega(\log n)$,
then the precomputed method will be asymptotically faster.

Rewriting algorithms in this way---to 
precompute some intermediate results that depend on constant (or slowly-varying) inputs---is 
highly non-trivial,
as it requires implementing more complex data structures and algorithms.  
In this paper, we present a splitting algorithm which does much of this task automatically.

@@

Our splitting algorithm scans \texttt{qss} for first-stage computations,
gathering them into \texttt{qs1}. Given \texttt{l}, this function performs these
computations and places the information needed by the subsequent function into a
boundary data structure.
%
In particular, \texttt{qs1} performs all recursive calls and evaluates all
instances of \texttt{part} (since it depends only on \texttt{l}). It
produces a boundary data structure that collects the results from these
recursive calls, tagged by the branch (\texttt{LT}, \texttt{EQ}, or \texttt{GT})
in which that call occurred. Since the recursive calls occur in two different
branches (\texttt{LT} and \texttt{GT}) the boundary structure is a binary tree.  
%
Lastly, it records \texttt{i} (the size of the left subtree) and \texttt{\#1 ht}
(the pivot/head of the list) in the boundary structure, because those
computations are \emph{held} for use in the second stage.
%
The final result is a binary search tree augmented with size information, and
whose keys are the pivots.

>>>>

We can attempt to improve on this algorithm by factoring out computations shared
between these calls to \texttt{qs}.
In particular, we can construct a binary search
tree out of \texttt{l}, at cost $\Theta(n\log{n})$, and then simply look for the \texttt{k}th
leftmost element of that tree. 
Doing lookups efficiently, in $\Theta(\log{n})$ time, 
requires one more innovation---augmenting the tree 
by storing at each node the size of its left subtree. 
Thus in total this approach has expected runtime $\Theta(n\log{n} + m\log{n})$.

Is this method, wherein we precompute a data structure, better than the direct $\Theta(mn)$ method? 
The answer depends on the relationship between the number of lookups $m$ and the size of the list $n$.
If $m$ is constant, then the direct method is superior for sufficiently large $n$.
This is because the precomputed method does unnecessary work 
sorting parts of the list where there are no query points.
However, if the number of queries grows with the size of the list, 
specficially in $\omega(\log n)$,
then the precomputed method will be asymptotically faster.

Rewriting algorithms in this way---to 
precompute some intermediate results that depend on constant (or infrequently-varying) inputs---is non-trivial,
as it requires implementing more complex data structures and algorithms.  
In this paper, we present a splitting algorithm which does much of this task automatically.

@@

Our splitting algorithm scans \texttt{qss} for first-stage computations,
gathering them into \texttt{qs1}. Given \texttt{l}, this function performs these
computations and places the information needed by the subsequent function into a
boundary data structure.
%
In particular, \texttt{qs1} performs all recursive calls and evaluates all
instances of \texttt{part} (since it depends only on \texttt{l}). It
produces a boundary data structure that collects the results from these
recursive calls, tagged by the branch (\texttt{LT}, \texttt{EQ}, or \texttt{GT})
in which that call occurred. Since the recursive calls occur in two different
branches (\texttt{LT} and \texttt{GT}) the boundary structure is a binary tree.  
%
Lastly, it records~\texttt{i} (the size of the left subtree) and \texttt{\#1 ht}
(the pivot/head of the list) in the boundary structure, because those
computations are \emph{held} for use in the second stage.
%
The final result is a binary search tree augmented with size information, and
whose keys are the pivots.

### 2016-01-08 paper/semantics.tex
mostly compression
small edits

The example code in this paper uses an extension of the formalized \lang.
In particular, it makes liberal use of $\rmint$s and various
functions on these, as well as a function $\pause$ which takes a $\curr\rmint$
to a $\fut\rmint$.%
\footnote{$\pause$ is definable in \lang\ given an inductive definition of
$\rmint$s. In practice, we provide both $\rmint$s and $\pause$ as primitives.
It is sensible to extend $\pause$ to all base types and to products and sums
thereof. This is related to the notion of {\em mobility} in \cite{murphy05} and
{\em stability} in \cite{krishnaswami13}.}

@@

We can also directly observe the reuse of first-stage computations across
multiple second-stage computations. To run many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
\begin{lstlisting}
2`let fun qs k = prev{`1`qss (`3`gr{`1`[7,4,2,5,9,...,3]`3`}`1`,next{`2`k`1`})2`}`
2`in (qs k1,...,qs km)`
\end{lstlisting}
To see that the partial application of \texttt{qss} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let fun qs k = (let val y = ... in y)
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $y$ is bound to a large nest of case statements open on $k$.

>>>>

The example code in this paper uses an extension of the formalized \lang.
In particular, it makes liberal use of $\rmint$s and various
functions on these, as well as a function $\pause$ which takes a $\curr\rmint$
to a $\fut\rmint$.%
\footnote{$\pause$ is definable in \lang\ given an inductive definition of
$\rmint$s. In practice, we provide both $\rmint$s and $\pause$ as primitives.
It is sensible to extend $\pause$ to all base types and to products and sums
thereof. This is related to the notion of {\em mobility} (\cite{murphy05}).}

@@

Moreover, the reuse of first-stage computations across
multiple second-stage computations can even be encoded within \lang. 
The following program runs many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list:
\begin{lstlisting}
2`prev{`1` let val list = `3`gr{`1`[7,4,2,5,9,...,3]`3`}`1` in 
next{ `2`let fun qs k = prev{`1`qss (list,next{`2`k`1`})2`}`
      2`in (qs k1,...,qs km)`1`}`2`}`
\end{lstlisting}
To see that the partial application of \texttt{qss} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let fun qs k = (let val y = ... in y)
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $y$ is bound to a large nest of case statements open on $k$.
